---
title: "Data Audit"
subtitle: "Completeness · Missingness · MAUP decisions · Crosswalk diagnostics"
date: last-modified
---

```{r setup, include=FALSE}
source(here::here("R", "setup.R"))
library(DT)

# Load completeness report from pipeline
completeness_path <- file.path(OUTPUT_DIR, "completeness_final.csv")
if (file.exists(completeness_path)) {
  comp <- readr::read_csv(completeness_path, show_col_types = FALSE)
} else {
  comp <- tibble::tibble(variable = character(), n_munis_with_data = integer(),
                         pct_munis = numeric(), source = character())
  warning("completeness_final.csv not found — run 04_assemble_panel.R first.")
}
```

## Data Sources & Availability {.tabset}

::: {.panel-tabset}

### Overview

```{r overview-table}
comp_summary <- comp |>
  dplyr::group_by(source) |>
  dplyr::summarise(
    n_vars     = dplyr::n(),
    pct_min    = round(min(pct_munis), 1),
    pct_median = round(median(pct_munis), 1),
    pct_max    = round(max(pct_munis), 1),
    n_zero     = sum(pct_munis == 0),
    n_full     = sum(pct_munis == 100),
    .groups = "drop"
  ) |>
  dplyr::arrange(desc(pct_median))

DT::datatable(
  comp_summary,
  colnames = c("Source", "# Variables", "Min %", "Median %", "Max %",
               "All-NA cols", "100% cols"),
  rownames = FALSE,
  options  = list(pageLength = 30, dom = "ftp"),
  caption  = glue::glue(
    "{nrow(comp)} variables across {nrow(comp_summary)} sources — ",
    "coverage = % of 170 PDET municipalities with ≥1 non-NA value"
  )
) |>
  DT::formatStyle(
    "pct_median",
    background = DT::styleColorBar(c(0, 100), "#c6efce"),
    backgroundSize = "98% 88%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  ) |>
  DT::formatStyle(
    "n_zero",
    color = DT::styleInterval(0, c("black", "#b2182b")),
    fontWeight = DT::styleInterval(0, c("normal", "bold"))
  )
```

### All Variables

Full searchable table of every variable in the panel with its completeness.

```{r all-vars-table}
DT::datatable(
  comp |>
    dplyr::arrange(source, desc(pct_munis)) |>
    dplyr::select(Source = source, Variable = variable,
                  `N munis with data` = n_munis_with_data,
                  `% coverage` = pct_munis),
  rownames = FALSE,
  filter   = "top",
  options  = list(
    pageLength = 25,
    dom        = "lfrtip",
    columnDefs = list(list(targets = 3, className = "dt-right"))
  ),
  caption = "One row per variable — filter by source or search by name"
) |>
  DT::formatStyle(
    "% coverage",
    background = DT::styleColorBar(c(0, 100), "#c6efce"),
    backgroundSize = "98% 88%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  ) |>
  DT::formatStyle(
    "% coverage",
    color = DT::styleInterval(c(1, 50, 90),
                               c("#b2182b", "#d6604d", "#555", "#1b7837"))
  )
```

### Conflict & Violence

CNMH event types, UCDP, VERDATA, police injuries, false positives.

```{r conflict-missingness}
conf_sources <- c("cnmh", "ucdp", "jep", "cerac", "daly", "aponte")
conf_comp <- dplyr::filter(comp, source %in% conf_sources |
                                  (category == "Violence & Conflict"))

DT::datatable(
  conf_comp |>
    dplyr::arrange(source, desc(pct_munis)) |>
    dplyr::select(Source = source, Variable = variable,
                  `N munis` = n_munis_with_data, `% coverage` = pct_munis),
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, dom = "lfrtip")
) |>
  DT::formatStyle(
    "% coverage",
    background = DT::styleColorBar(c(0, 100), "#fadbd8"),
    backgroundSize = "98% 88%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

### Peacebuilding & Justice

SIIPO, Pilar 1 investment, SLJ, Casas de Justicia, Conciliadores, PDET initiatives.

```{r peace-missingness}
peace_sources <- c("art", "apc", "pdet", "slj", "casjust", "concil", "urt")
peace_comp <- dplyr::filter(comp, source %in% peace_sources |
                                   category %in% c("Peacebuilding & Implementation",
                                                    "Justice & Dispute Resolution"))

DT::datatable(
  peace_comp |>
    dplyr::arrange(source, desc(pct_munis)) |>
    dplyr::select(Source = source, Variable = variable,
                  `N munis` = n_munis_with_data, `% coverage` = pct_munis),
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, dom = "lfrtip")
) |>
  DT::formatStyle(
    "% coverage",
    background = DT::styleColorBar(c(0, 100), "#d4edda"),
    backgroundSize = "98% 88%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

### Controls & Structural

DNP Panel General, DNP Fiscal, GINI, PRIO-GRID, Territorial.

```{r controls-missingness}
ctrl_sources <- c("dnp", "igac", "prio", "ant")
ctrl_comp <- dplyr::filter(comp, source %in% ctrl_sources |
                                  category %in% c("Socioeconomic & Structural",
                                                   "Geographic & Territorial"))

DT::datatable(
  ctrl_comp |>
    dplyr::arrange(source, desc(pct_munis)) |>
    dplyr::select(Source = source, Variable = variable,
                  `N munis` = n_munis_with_data, `% coverage` = pct_munis),
  rownames = FALSE,
  filter   = "top",
  options  = list(pageLength = 20, dom = "lfrtip")
) |>
  DT::formatStyle(
    "% coverage",
    background = DT::styleColorBar(c(0, 100), "#dce9f5"),
    backgroundSize = "98% 88%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

:::

---

## MAUP Decisions Log

```{r maup-log}
maup_log <- tibble::tribble(
  ~Decision,                      ~Choice,                                  ~Rationale,                                                                                          ~Script,             ~Date,
  "Primary spatial unit",         "Municipality polygon (GADM L2)",         "Unit of policy intervention (PDET designation is municipal); standard in Colombian conflict lit.", "01_build_geometry.R", "2026-03-14",
  "CRS for area calculations",    "MAGNA-SIRGAS / EPSG:3116",               "Colombian national projected CRS; minimizes area distortion.",                                      "01_build_geometry.R", "2026-03-14",
  "UCDP point-to-polygon join",   "st_within + nearest-feature fallback",   "st_nearest_feature as fallback when point falls outside polygon (coastal/island municipalities).", "03_ucdp_maup.R",      "2026-03-14",
  "UCDP boundary buffer",         "5 km from internal boundaries",          "Events within 5 km flagged; sensitivity tested at 1, 2, 5, 10, 20 km.",                            "03_ucdp_maup.R",      "2026-03-16",
  "Spatial weights (primary)",    "Queen contiguity, W-standardized",       "Standard choice; compared against rook and knn4.",                                                  "03_ucdp_maup.R",      "2026-03-16",
  "DANE code merge strategy",     "Exact DANE code only (no fuzzy)",        "All joins by exact 5-digit DANE code; name lookup only for geometry attachment.",                  "02_build_tabular_panel.R","2026-04-04",
  "Panel format",                 "Long (170 munis × 25 years = 9,251 rows)","One row per municipality × year; static variables broadcast across all years.",                  "04_assemble_panel.R", "2026-04-04",
  "VERDATA aggregation",          "Mean of 100 replicas + 95% CI cols",     "Accounts for uncertainty in VERDATA posterior; CI available for sensitivity analysis.",            "spatial/spatial_joins.R","2026-03-14",
  "PDET filter for maps",         "All Colombia outline + PDET data",       "Non-PDET shown as gray polygon outline; data shown only for PDET (n=170).",                        "R/setup.R",           "2026-03-20",
  "Spatial weights for PDET-only","Queen, PDET polygons only",              "For regime models restricted to n=170; built on PDET subset to avoid ghost neighbors.",            "panel3_regression.qmd","TODO"
)

DT::datatable(
  maup_log,
  rownames = FALSE,
  options  = list(pageLength = 15, dom = "ftp"),
  caption  = "Spatial modeling decisions log"
)
```

---

## Crosswalk Diagnostics

```{r crosswalk-diag}
# Show match rates from pipeline output
if (nrow(comp) > 0) {
  xwalk <- comp |>
    dplyr::group_by(source) |>
    dplyr::summarise(
      munis_at_100pct = sum(pct_munis == 100),
      munis_any_data  = round(max(pct_munis), 1),
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(munis_any_data))

  DT::datatable(
    xwalk,
    colnames = c("Data Source", "Variables at 100%", "Best coverage (%)"),
    rownames = FALSE,
    options  = list(pageLength = 20, dom = "ft")
  )
} else {
  cat("> Run 04_assemble_panel.R to populate completeness diagnostics.\n")
}
```

---

## Gap Analysis vs. Proposal

```{r gap-analysis}
gaps <- tibble::tribble(
  ~`Dataset`,                       ~`Status`,    ~`Notes`,
  "CNMH Casos (11 event types)",    "✅ In panel", "V_*__cnmh; 161/170 munis",
  "UCDP GED v25.1",                 "✅ In panel", "V_conflict_events/battle_deaths__ucdp; MAUP outputs in output/",
  "VERDATA Homicidio",              "✅ In panel", "V_homicides_mean__jep; 100 replicas mean + CI",
  "False Positives / Extrajudicial","✅ In panel", "V_false_positive/extrajudicial__cerac; 169/170 munis",
  "Daly Paramilitary (to 2015)",    "✅ In panel", "V_criminal_pact/political_violence__daly; 170/170 munis",
  "Aponte González Control Data",   "✅ In panel", "V_territorial_control/coca__aponte; 76/170 munis (armed group presence zones)",
  "SIIPO Implementation Index",     "✅ In panel", "P_impl_*__art; pivoted wider (18 pilars); 170/170 munis",
  "Pilar 1 Investment",             "✅ In panel", "P_pilar1_investment_cop__art; 146/170 munis",
  "ART Obras Terminadas",           "✅ In panel", "P_n_projects/investment_total__art; muni×year construction projects",
  "APC Cooperación Internacional",  "✅ In panel", "P_aid_usd/n_interventions__apc; 10,501/10,525 rows matched via divipola",
  "PDET Iniciativas",               "✅ In panel", "P_n_initiatives_*__pdet; static counts by pilar",
  "CNMH Memory Initiatives",        "✅ In panel", "P_n_memory_initiatives__cnmh",
  "SLJ / Casas de Justicia",        "✅ In panel", "J_n_justice_rooms/slj_present__slj; J_n_cases_processed__casjust",
  "Conciliadores en Equidad",       "✅ In panel", "J_n_conciliadores/conciliation_requests__concil",
  "URT Land Restitution",           "✅ In panel", "J_n_restitution_sentences__urt",
  "DNP Panel Caracteristicas",      "✅ In panel", "S_pop/gdp/nbi/ipm/poverty__dnp; 170/170 munis",
  "DNP Fiscal / TerriData",         "✅ In panel", "S_income_total_pc/tax_income_pc__dnp; per-capita only",
  "GINI Municipal (IGAC 2023)",     "✅ In panel", "S_gini_land__igac; 161/170 munis",
  "PRIO-GRID (static + time-var)",  "✅ In panel", "G_terrain/nightlight/forest__prio",
  "Territorial (ANT)",              "✅ In panel", "G_indigenous/afro/peasant_pct__ant",
  "DNP Capacidad Fiscal",           "⚠️ 0% values", "File loaded but values all NA — investigate DNP fiscal format",
  "Lesiones Policiales",            "⚠️ 0% values", "Source file empty or key column mismatch — dropped from panel",
  "Rutas de Implementacion Activa", "❌ Pending",  "No DANE code column; needs manual inspection",
  "PNIS Matriz",                    "❌ Pending",  "No DANE code column; wide format needs restructuring",
  "AvanceMensualRIA",               "❌ Missing",  "File not in Dropbox — check sync",
  "Veedurías Ciudadanas",           "❌ Pending",  "33 dept xlsx; name-based join needed",
  "Elecciones",                     "❌ Pending",  "Electoral results; key column TBD"
)

DT::datatable(
  gaps,
  rownames = FALSE,
  escape   = FALSE,
  options  = list(pageLength = 30, dom = "ftp"),
  caption  = "Dataset tracking — pipeline status as of 2026-04-04"
) |>
  DT::formatStyle(
    "Status",
    color = DT::styleEqual(
      c("✅ In panel", "⚠️ 0% values", "❌ Pending", "❌ Missing"),
      c("#1b7837",    "#c8923a",       "#b2182b",    "#555555")
    )
  )
```
