---
title: "Codebook"
subtitle: "Variable dictionary for the PDET municipality panel"
date: last-modified
---

```{r setup, include=FALSE}
source(here::here("R", "setup.R"))
library(DT)

# Load completeness report
comp_path <- file.path(OUTPUT_DIR, "completeness_final.csv")
if (file.exists(comp_path)) {
  comp <- readr::read_csv(comp_path, show_col_types = FALSE)
} else {
  comp <- tibble::tibble(variable = character(), n_munis_with_data = integer(),
                         pct_munis = numeric(), source = character(),
                         category = character())
  warning("completeness_final.csv not found — run 04_assemble_panel.R first.")
}
```

## Panel Overview

```{r panel-overview}
n_munis <- dplyr::n_distinct(panel$dane_code)
n_years <- dplyr::n_distinct(panel$year)
yr_range <- range(panel$year, na.rm = TRUE)
data_vars <- setdiff(names(panel), c("dane_code", "year", "muni_name", "dept_name", "subregion"))

overview <- tibble::tribble(
  ~Metric,                ~Value,
  "Municipalities (PDET)", as.character(n_munis),
  "Years",                 sprintf("%d–%d (%d years)", yr_range[1], yr_range[2], n_years),
  "Total rows",            format(nrow(panel), big.mark = ","),
  "Data variables",        as.character(length(data_vars)),
  "Sources",               as.character(dplyr::n_distinct(comp$source))
)

knitr::kable(overview, col.names = c("", ""), align = "lr")
```

## Naming Convention

All variables follow the pattern: **`{Category}_{name}__{source}`**

| Code | Category | Example |
|---|---|---|
| `V` | Violence & Conflict | `V_massacre_cases__cnmh` |
| `P` | Peacebuilding & Implementation | `P_impl_education__art` |
| `J` | Justice & Dispute Resolution | `J_n_conciliadores__concil` |
| `S` | Socioeconomic & Structural | `S_nbi__dnp` |
| `G` | Geographic & Territorial | `G_nightlight__prio` |

### Source Abbreviations

| Abbrev | Full Name |
|---|---|
| `cnmh` | Centro Nacional de Memoria Histórica |
| `ucdp` | Uppsala Conflict Data Program (GED v25.1) |
| `jep` | JEP / VERDATA statistical homicide estimates |
| `cerac` | CERAC False Positives dataset |
| `daly` | Daly paramilitary-mayor pacts dataset |
| `aponte` | Aponte González armed group control data |
| `art` | Agencia de Renovación del Territorio (SIIPO, Pilar 1, Obras) |
| `apc` | Agencia Presidencial de Cooperación Internacional |
| `pdet` | PDET peace initiative counts |
| `slj` | Sistemas Locales de Justicia |
| `casjust` | Casas de Justicia |
| `concil` | Conciliadores en Equidad / Solicitudes de Conciliación |
| `urt` | Unidad de Restitución de Tierras |
| `dnp` | Departamento Nacional de Planeación (Panel General + Fiscal) |
| `igac` | IGAC (GINI municipal + Gestores Catastrales) |
| `prio` | Peace Research Institute Oslo (PRIO-GRID) |
| `ant` | Agencia Nacional de Tierras (Resguardos, Zonas Reserva, Consejos) |

---

## Full Variable Dictionary {.tabset}

::: {.panel-tabset}

### Searchable Table

```{r codebook-table}
if (nrow(comp) > 0) {
  DT::datatable(
    comp |>
      dplyr::select(
        Category    = category,
        Source      = source,
        Variable    = variable,
        `N munis`   = n_munis_with_data,
        `% coverage` = pct_munis
      ) |>
      dplyr::arrange(Category, Source, dplyr::desc(`% coverage`)),
    rownames = FALSE,
    filter   = "top",
    escape   = FALSE,
    options  = list(
      pageLength = 40,
      dom        = "lfrtip",
      columnDefs = list(
        list(targets = c(3, 4), className = "dt-right")
      )
    ),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-size: 0.9em;",
      sprintf("Full codebook: %d variables, %d sources, %d categories",
              nrow(comp), dplyr::n_distinct(comp$source), dplyr::n_distinct(comp$category))
    )
  ) |>
    DT::formatStyle(
      "% coverage",
      background = DT::styleColorBar(c(0, 100), "#c6efce"),
      backgroundSize = "98% 88%",
      backgroundRepeat = "no-repeat",
      backgroundPosition = "center"
    ) |>
    DT::formatStyle(
      "% coverage",
      color = DT::styleInterval(c(1, 50, 90),
                                 c("#b2182b", "#d6604d", "#555", "#1b7837"))
    ) |>
    DT::formatStyle(
      "Category",
      color = DT::styleEqual(
        c("Violence & Conflict", "Peacebuilding & Implementation",
          "Justice & Dispute Resolution", "Socioeconomic & Structural",
          "Geographic & Territorial"),
        c("#b2182b", "#1b7837", "#2166ac", "#8c510a", "#542788")
      ),
      fontWeight = "bold"
    )
} else {
  cat("> No data available. Run the pipeline first.\n")
}
```

### By Category

```{r by-category}
if (nrow(comp) > 0) {
  cat_summary <- comp |>
    dplyr::group_by(category) |>
    dplyr::summarise(
      `N variables`     = dplyr::n(),
      `N sources`       = dplyr::n_distinct(source),
      `Min coverage`    = round(min(pct_munis), 1),
      `Median coverage` = round(median(pct_munis), 1),
      `Max coverage`    = round(max(pct_munis), 1),
      .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(`N variables`))

  DT::datatable(
    cat_summary,
    rownames = FALSE,
    options  = list(pageLength = 10, dom = "t"),
    caption  = "Summary by thematic category"
  ) |>
    DT::formatStyle(
      "Median coverage",
      background = DT::styleColorBar(c(0, 100), "#c6efce"),
      backgroundSize = "98% 88%",
      backgroundRepeat = "no-repeat",
      backgroundPosition = "center"
    )
}
```

### By Source

```{r by-source}
if (nrow(comp) > 0) {
  src_summary <- comp |>
    dplyr::group_by(source, category) |>
    dplyr::summarise(
      `N vars`          = dplyr::n(),
      `Median coverage` = round(median(pct_munis), 1),
      .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(`Median coverage`))

  DT::datatable(
    src_summary,
    rownames = FALSE,
    options  = list(pageLength = 20, dom = "ft"),
    caption  = "Coverage summary by data source"
  ) |>
    DT::formatStyle(
      "Median coverage",
      background = DT::styleColorBar(c(0, 100), "#c6efce"),
      backgroundSize = "98% 88%",
      backgroundRepeat = "no-repeat",
      backgroundPosition = "center"
    )
}
```

:::

---

## Data Sources

| Source | Institution | Period | Category |
|---|---|---|---|
| DNP Panel General | Dept. Nacional de Planeación | 2000–2023 | S |
| DNP Fiscal | DNP TerriData | 2000–2023 | S |
| CNMH Casos | Centro Nacional de Memoria Histórica | 1958–2023 | V |
| UCDP GED v25.1 | Uppsala Conflict Data Program | 1989–2024 | V |
| VERDATA / JEP | JEP Statistical Commission | 1985–2018 | V |
| False Positives | CERAC | 2000–2010 | V |
| Daly Paramilitary | Academic dataset | 1981–2019 | V |
| Aponte González | Academic dataset | 2000–2012 | V |
| SIIPO | Agencia de Renovación del Territorio | 2017–2021 | P |
| APC Cooperación | Agencia Presidencial de Cooperación | 2018–2025 | P |
| ART Obras | Agencia de Renovación del Territorio | 2012–2025 | P |
| Pilar 1 Investment | ART / DNP | 2016–2025 | P |
| PDET Initiatives | ART | Static | P |
| PRIO-GRID | Peace Research Institute Oslo | Static + 1990–2020 | G |
| Territorial / ANT | Agencia Nacional de Tierras | Static | G |
| GINI | IGAC 2023 | Static | S |
| SLJ / Casas de Justicia | Min. Justicia | Static + year-varying | J |
| Conciliadores | Min. Justicia | 1993–2024 | J |
| URT | Unidad de Restitución de Tierras | Static | J |
