---
title: "Variable Explorer"
subtitle: "Interactive choropleth maps · Moran's I · LISA clusters"
date: last-modified
---

```{r setup, include=FALSE}
source(here::here("R", "setup.R"))
library(DT)
library(spdep)

# ── Which variables to display on the website ─────────────────────────────────
# Toggle TRUE/FALSE to show/hide variables in the explorer.
# Edit this list to control what appears in the dropdown.

EXPLORER_VARS <- c(
  # ── V: Violence & Conflict ──
  # Set TRUE to show on website, FALSE to hide. Toggle as needed.
  "V_selective_killing_cases__cnmh"        = TRUE,
  "V_massacre_cases__cnmh"                 = TRUE,
  "V_forced_disappearance_cases__cnmh"     = TRUE,
  "V_kidnapping_cases__cnmh"               = TRUE,
  "V_sexual_violence_cases__cnmh"          = TRUE,
  "V_landmine_cases__cnmh"                 = TRUE,
  "V_child_recruitment_cases__cnmh"        = TRUE,
  "V_conflict_events_all__ucdp"            = TRUE,
  "V_battle_deaths_all__ucdp"              = TRUE,
  "V_homicides_mean__jep"                  = TRUE,
  "V_false_positive_cases__cerac"          = TRUE,
  "V_extrajudicial_executions__cerac"      = TRUE,
  "V_territorial_control_status__aponte"   = TRUE,
  "V_coca_cultivation__aponte"             = TRUE,
  "V_criminal_pact__daly"                  = TRUE,
  "V_political_violence_index__daly"       = TRUE,

  # ── P: Peacebuilding & Implementation ──
  "P_impl_mean_all_pilars__art"            = TRUE,
  "P_impl_education__art"                  = TRUE,
  "P_impl_reconciliation__art"             = TRUE,
  "P_impl_security_guarantees__art"        = TRUE,
  "P_pilar1_investment_cop__art"           = TRUE,
  "P_n_projects__art"                      = TRUE,
  "P_investment_total_cop__art"            = TRUE,
  "P_aid_usd__apc"                         = TRUE,
  "P_n_interventions__apc"                 = TRUE,
  "P_n_initiatives_total__pdet"            = TRUE,
  "P_n_memory_initiatives__cnmh"           = TRUE,

  # ── J: Justice & Dispute Resolution ──
  "J_n_justice_rooms__slj"                 = TRUE,
  "J_slj_present__slj"                     = TRUE,
  "J_n_cases_processed__casjust"           = TRUE,
  "J_n_conciliadores__concil"              = TRUE,
  "J_n_conciliation_requests__concil"      = TRUE,
  "J_n_restitution_sentences__urt"         = TRUE,

  # ── S: Socioeconomic ──
  "S_pop_total__dnp"                       = TRUE,
  "S_gdp_percapita_constant__dnp"          = TRUE,
  "S_nbi__dnp"                             = TRUE,
  "S_ipm__dnp"                             = TRUE,
  "S_poverty_headcount__dnp"               = TRUE,
  "S_gini_land__igac"                      = TRUE,
  "S_income_total_pc__dnp"                 = TRUE,
  "S_tax_income_pc__dnp"                   = TRUE,
  "S_ethnic_minority_share__dnp"           = TRUE,

  # ── G: Geographic ──
  "G_terrain_ruggedness__prio"             = TRUE,
  "G_nightlight__prio"                     = TRUE,
  "G_forest_share__prio"                   = TRUE,
  "G_indigenous_pct_covered__ant"          = TRUE,
  "G_afro_pct_covered__ant"               = TRUE,
  "G_peasant_pct_covered__ant"             = TRUE
)

# Filter to enabled variables that exist in the panel
enabled_vars <- names(EXPLORER_VARS)[EXPLORER_VARS]
enabled_vars <- intersect(enabled_vars, names(panel))

# ── Category definitions ──
CAT_INFO <- list(
  V = list(label = "Violence & Conflict",          color = "#b2182b", palette = "YlOrRd"),
  P = list(label = "Peacebuilding & Implementation", color = "#1b7837", palette = "YlGn"),
  J = list(label = "Justice & Dispute Resolution",  color = "#2166ac", palette = "Blues"),
  S = list(label = "Socioeconomic & Structural",    color = "#8c510a", palette = "YlOrBr"),
  G = list(label = "Geographic & Territorial",      color = "#542788", palette = "PuBuGn")
)

# Split enabled vars by category prefix
vars_by_cat <- split(enabled_vars,
                     sub("^([VPJSG])_.*", "\\1", enabled_vars))

# ── Helper: build ONE static faceted ggplot map per category ──
# Replaces the prior multi-layer leaflet (which embedded full PDET GeoJSON
# per variable and overflowed V8's ~512MB string buffer during render).
build_cat_map <- function(cat_vars, viridis_option = "C") {
  if (length(cat_vars) == 0) return(NULL)

  default_year <- max(panel$year)
  numeric_vars <- cat_vars[sapply(panel[cat_vars], is.numeric)]
  if (length(numeric_vars) == 0) return(NULL)

  cross <- panel |>
    dplyr::filter(year == default_year) |>
    dplyr::select(dane_code, dplyr::all_of(numeric_vars))

  map_sf <- pdet_munis |>
    dplyr::left_join(cross, by = "dane_code")

  # Drop variables that are all-NA at the chosen year
  keep_vars <- numeric_vars[vapply(numeric_vars,
                                   function(v) any(!is.na(map_sf[[v]])),
                                   logical(1))]
  if (length(keep_vars) == 0) return(NULL)

  # Long-format sf: one row per (muni × variable)
  plot_list <- lapply(keep_vars, function(v) {
    label <- gsub("__.*", "", gsub("^[VPJSG]_", "", v))
    map_sf |>
      dplyr::transmute(
        dane_code,
        muni_name,
        value     = .data[[v]],
        var_label = label
      )
  })
  plot_sf <- dplyr::bind_rows(plot_list) |> sf::st_as_sf()

  # log1p transform for display (handles skew); preserve sign via pmax(0)
  plot_sf <- plot_sf |>
    dplyr::mutate(value_display = log1p(pmax(value, 0, na.rm = TRUE)))

  ncols <- min(3L, length(keep_vars))

  ggplot2::ggplot(plot_sf) +
    ggplot2::geom_sf(ggplot2::aes(fill = value_display),
                     color = "white", linewidth = 0.15) +
    ggplot2::scale_fill_viridis_c(
      option   = viridis_option,
      na.value = "#e0e0e0",
      name     = "log(1+x)",
      guide    = ggplot2::guide_colorbar(barwidth = 6, barheight = 0.5,
                                         title.position = "top")
    ) +
    ggplot2::facet_wrap(~var_label, ncol = ncols) +
    ggplot2::theme_void(base_size = 9) +
    ggplot2::theme(
      legend.position  = "bottom",
      strip.text       = ggplot2::element_text(size = 7, face = "bold",
                                                margin = ggplot2::margin(b = 3)),
      plot.margin      = ggplot2::margin(4, 4, 4, 4)
    ) +
    ggplot2::labs(caption = paste0("Year: ", default_year))
}
```


## Interactive Choropleth Maps

::: {.callout-tip}
**How to use this explorer**

1. Choose a **category tab** (Violence, Peacebuilding, Justice, Socioeconomic, Geographic, or Year Comparison).
2. Each tab shows all enabled variables in that category as a **faceted map** (one small-multiple panel per variable).
3. Gray polygons = no data available for that variable in that municipality.
4. Maps show the **most recent year** with data for each variable. The Variable Atlas shows the all-years mean instead.
5. Values are shown on a log(1+x) scale to handle right-skewed count variables.

The Year Comparison tab shows UCDP conflict events at five-year intervals to visualize the conflict trajectory from 2005 to 2024.
:::

---

::: {.panel-tabset}

### Violence & Conflict

```{r map-violence}
#| fig-height: 10
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis) && nrow(panel) > 0 && "V" %in% names(vars_by_cat)) {
  build_cat_map(vars_by_cat[["V"]], "C")
}
```

### Peacebuilding

```{r map-peace}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis) && nrow(panel) > 0 && "P" %in% names(vars_by_cat)) {
  build_cat_map(vars_by_cat[["P"]], "D")
}
```

### Justice & Dispute

```{r map-justice}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis) && nrow(panel) > 0 && "J" %in% names(vars_by_cat)) {
  build_cat_map(vars_by_cat[["J"]], "B")
}
```

### Socioeconomic

```{r map-socio}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis) && nrow(panel) > 0 && "S" %in% names(vars_by_cat)) {
  build_cat_map(vars_by_cat[["S"]], "A")
}
```

### Geographic

```{r map-geo}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis) && nrow(panel) > 0 && "G" %in% names(vars_by_cat)) {
  build_cat_map(vars_by_cat[["G"]], "E")
}
```

### Year Comparison

```{r year-panels}
#| fig-height: 6

if (!is.null(pdet_munis) && "V_conflict_events_all__ucdp" %in% names(panel)) {
  key_years <- c(2005, 2010, 2015, 2020, 2024)
  key_var <- "V_conflict_events_all__ucdp"

  year_data <- panel |>
    dplyr::filter(year %in% key_years) |>
    dplyr::select(dane_code, year, value = all_of(key_var))

  map_data <- pdet_munis |>
    dplyr::left_join(year_data, by = "dane_code")

  ggplot(map_data) +
    geom_sf(aes(fill = value), color = "white", size = 0.1) +
    scale_fill_viridis_c(option = "inferno", na.value = "#d9d9d9",
                          name = gsub("__.*", "", gsub("^V_", "", key_var))) +
    facet_wrap(~year, nrow = 1) +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      legend.position = "bottom"
    ) +
    labs(title = paste("UCDP Conflict Events:", paste(key_years, collapse = ", ")))
}
```

:::

---

## Spatial Autocorrelation: Justice & Dispute Variables {.tabset}

Global Moran's I and LISA cluster maps for all **J_** (Justice & Dispute Resolution) variables. These diagnostics assess whether dispute resolution capacity clusters spatially --- a prerequisite for constructing a latent dispute resolution variable.

::: {.panel-tabset}

### Global Moran's I

```{r moran-global}
if (!is.null(pdet_munis) && nrow(panel) > 0) {

  # Build spatial weights (Queen contiguity on PDET polygons)
  pdet_sp <- sf::as_Spatial(pdet_munis)
  nb <- spdep::poly2nb(pdet_sp, queen = TRUE)
  lw <- spdep::nb2listw(nb, style = "W", zero.policy = TRUE)

  # Get J_ variables + key V_ and P_ variables for comparison
  dispute_vars <- grep("^J_", names(panel), value = TRUE)
  comparison_vars <- c(
    "V_selective_killing_cases__cnmh",
    "V_conflict_events_all__ucdp",
    "V_homicides_mean__jep",
    "P_impl_mean_all_pilars__art",
    "P_aid_usd__apc",
    "S_nbi__dnp"
  )
  test_vars <- intersect(c(dispute_vars, comparison_vars), names(panel))

  # Cross-section: aggregate across all years (mean per municipality)
  cross_mean <- panel |>
    dplyr::group_by(dane_code) |>
    dplyr::summarise(dplyr::across(dplyr::all_of(test_vars),
                                    ~ mean(.x, na.rm = TRUE)),
                     .groups = "drop")

  # Join to geometry (preserve order)
  geo_cross <- pdet_munis |>
    dplyr::left_join(cross_mean, by = "dane_code")

  # Compute Moran's I for each variable
  moran_results <- purrr::map_dfr(test_vars, function(v) {
    x <- geo_cross[[v]]
    if (sum(!is.na(x)) < 30) {
      return(tibble::tibble(variable = v, morans_I = NA_real_,
                            p_value = NA_real_, z_score = NA_real_))
    }
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    if (sd(x, na.rm = TRUE) == 0 || is.na(sd(x, na.rm = TRUE))) {
      return(tibble::tibble(variable = v, morans_I = NA_real_,
                            p_value = NA_real_, z_score = NA_real_))
    }

    mt <- tryCatch(
      spdep::moran.test(x, lw, zero.policy = TRUE),
      error = function(e) NULL
    )
    if (is.null(mt)) {
      return(tibble::tibble(variable = v, morans_I = NA_real_,
                            p_value = NA_real_, z_score = NA_real_))
    }
    tibble::tibble(
      variable = v,
      morans_I = round(mt$estimate["Moran I statistic"], 4),
      p_value  = round(mt$p.value, 6),
      z_score  = round(mt$statistic, 2)
    )
  }) |>
    dplyr::mutate(
      category = dplyr::case_when(
        stringr::str_starts(variable, "J_") ~ "Justice & Dispute",
        stringr::str_starts(variable, "V_") ~ "Violence (comparison)",
        stringr::str_starts(variable, "P_") ~ "Peace (comparison)",
        stringr::str_starts(variable, "S_") ~ "Socioeconomic (comparison)",
        TRUE ~ "Other"
      ),
      significant = ifelse(p_value < 0.05, "***", "")
    ) |>
    dplyr::arrange(dplyr::desc(morans_I))

  DT::datatable(
    moran_results,
    rownames = FALSE,
    options  = list(pageLength = 20, dom = "ft"),
    caption  = "Global Moran's I — Queen contiguity, W-standardized. Mean across all years."
  ) |>
    DT::formatStyle(
      "category",
      color = DT::styleEqual(
        c("Justice & Dispute", "Violence (comparison)", "Peace (comparison)", "Socioeconomic (comparison)"),
        c("#2166ac", "#b2182b", "#1b7837", "#8c510a")
      ),
      fontWeight = "bold"
    ) |>
    DT::formatStyle(
      "significant",
      color = "red",
      fontWeight = "bold"
    )
}
```

### LISA Cluster Maps

```{r lisa-maps}
#| fig-height: 10
#| fig-width: 9
#| out-width: "100%"

if (!is.null(pdet_munis) && nrow(panel) > 0) {

  # LISA for each J_ variable — build a single long-format sf and facet_wrap
  # (avoids patchwork-over-many-ggplots and keeps HTML small).
  lisa_vars <- intersect(dispute_vars, names(geo_cross))

  cluster_levels <- c("High-High", "Low-Low", "High-Low",
                      "Low-High", "Not significant")
  cluster_colors <- c(
    "High-High"       = "#d7191c",
    "Low-Low"         = "#2c7bb6",
    "High-Low"        = "#fdae61",
    "Low-High"        = "#abd9e9",
    "Not significant" = "#e0e0e0"
  )

  if (length(lisa_vars) > 0) {
    panels <- purrr::map(lisa_vars, function(v) {
      x <- geo_cross[[v]]
      if (sum(!is.na(x)) < 20) return(NULL)
      x[is.na(x)] <- mean(x, na.rm = TRUE)
      if (sd(x, na.rm = TRUE) == 0 || is.na(sd(x, na.rm = TRUE))) return(NULL)

      lm <- tryCatch(
        spdep::localmoran(x, lw, zero.policy = TRUE),
        error = function(e) NULL
      )
      if (is.null(lm)) return(NULL)

      xz <- scale(x)[, 1]
      if (any(!is.finite(xz))) return(NULL)
      lag_xz <- spdep::lag.listw(lw, xz, zero.policy = TRUE)
      p_local <- lm[, "Pr(z != E(Ii))"]

      clean_name <- gsub("^J_", "", gsub("__.*", "", v))
      mi_val <- moran_results$morans_I[moran_results$variable == v]
      panel_label <- sprintf("%s (I=%.3f)", clean_name,
                             if (length(mi_val)) mi_val else NA_real_)

      geo_cross |>
        dplyr::transmute(
          dane_code,
          var_label = panel_label,
          cluster   = factor(dplyr::case_when(
            p_local > 0.05       ~ "Not significant",
            xz > 0 & lag_xz > 0  ~ "High-High",
            xz < 0 & lag_xz < 0  ~ "Low-Low",
            xz > 0 & lag_xz < 0  ~ "High-Low",
            xz < 0 & lag_xz > 0  ~ "Low-High",
            TRUE                  ~ "Not significant"
          ), levels = cluster_levels)
        )
    })

    panels <- purrr::compact(panels)
    if (length(panels) > 0) {
      lisa_sf <- dplyr::bind_rows(panels) |> sf::st_as_sf()
      ncols <- min(3L, length(unique(lisa_sf$var_label)))

      ggplot2::ggplot(lisa_sf) +
        ggplot2::geom_sf(ggplot2::aes(fill = cluster),
                         color = "white", linewidth = 0.15) +
        ggplot2::scale_fill_manual(values = cluster_colors,
                                   name   = "LISA Cluster",
                                   drop   = FALSE) +
        ggplot2::facet_wrap(~var_label, ncol = ncols) +
        ggplot2::theme_void(base_size = 9) +
        ggplot2::theme(
          legend.position = "bottom",
          strip.text      = ggplot2::element_text(size = 7, face = "bold",
                                                  margin = ggplot2::margin(b = 3)),
          plot.margin     = ggplot2::margin(4, 4, 4, 4)
        )
    } else {
      cat("> No J_ variables with sufficient data for LISA.\n")
    }
  }
}
```

### Moran Scatterplots

```{r moran-scatter}
#| fig-height: 8

if (!is.null(pdet_munis) && nrow(panel) > 0) {

  scatter_plots <- purrr::map(lisa_vars, function(v) {
    x <- geo_cross[[v]]
    if (sum(!is.na(x)) < 20) return(NULL)
    x[is.na(x)] <- mean(x, na.rm = TRUE)
    if (sd(x, na.rm = TRUE) == 0 || is.na(sd(x, na.rm = TRUE))) return(NULL)

    xz <- scale(x)[, 1]
    if (any(!is.finite(xz))) return(NULL)
    lag_xz <- spdep::lag.listw(lw, xz, zero.policy = TRUE)

    clean_name <- gsub("^J_", "", gsub("__.*", "", v))

    df <- data.frame(x = xz, lag_x = lag_xz)
    ggplot(df, aes(x = x, y = lag_x)) +
      geom_point(alpha = 0.6, size = 2, color = "#2166ac") +
      geom_smooth(method = "lm", se = FALSE, color = "#b2182b", linewidth = 1) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      theme_minimal() +
      labs(title = clean_name,
           x = "Standardized value",
           y = "Spatial lag") +
      annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.5,
               label = sprintf("I = %.3f",
                                moran_results$morans_I[moran_results$variable == v]),
               size = 4, color = "#b2182b")
  })

  scatter_plots <- purrr::compact(scatter_plots)
  if (length(scatter_plots) > 0) {
    patchwork::wrap_plots(scatter_plots, ncol = 2)
  }
}
```

:::

---

::: {.callout-note}
## Research context
The spatial diagnostics above test whether dispute resolution infrastructure (justice rooms, conciliators, restitution sentences) clusters geographically. **Positive spatial autocorrelation** (high Moran's I, significant High-High and Low-Low clusters) would support constructing a latent dispute resolution capacity variable using spatial factor analysis methods.
:::
