---
title: "Variable Atlas"
subtitle: "Choropleth per category · Global Moran's I · Missingness map"
date: last-modified
---

```{r setup, include=FALSE}
source(here::here("R", "setup.R"))
library(DT)

# ── Build cross-section: mean per municipality across all years ──
data_vars    <- setdiff(names(panel), c("dane_code", "year"))
numeric_vars <- data_vars[sapply(panel[data_vars], is.numeric)]

cross_mean <- panel |>
  dplyr::group_by(dane_code) |>
  dplyr::summarise(dplyr::across(dplyr::all_of(numeric_vars),
                                  ~ mean(.x, na.rm = TRUE)),
                   .groups = "drop")

# ── Spatial weights (PDET only, queen contiguity) ──
lw_queen <- NULL
if (!is.null(pdet_munis)) {
  nb_queen <- spdep::poly2nb(sf::as_Spatial(pdet_munis), queen = TRUE)
  lw_queen <- spdep::nb2listw(nb_queen, style = "W", zero.policy = TRUE)
}

compute_morans <- function(vals) {
  if (is.null(lw_queen)) return(c(I = NA_real_, p = NA_real_))
  vals[is.na(vals)] <- 0
  if (var(vals) == 0) return(c(I = NA_real_, p = NA_real_))
  mi <- tryCatch(
    suppressWarnings(spdep::moran.test(vals, lw_queen, zero.policy = TRUE)),
    error = function(e) NULL
  )
  if (is.null(mi)) return(c(I = NA_real_, p = NA_real_))
  c(I = unname(mi$estimate["Moran I statistic"]), p = mi$p.value)
}

# ── Align cross-section to geometry order ──
geo_cross <- pdet_munis |>
  dplyr::left_join(cross_mean, by = "dane_code")

# ── Key variables per category ──
ATLAS_VARS <- list(
  V = c("V_selective_killing_cases__cnmh", "V_massacre_cases__cnmh",
        "V_forced_disappearance_cases__cnmh", "V_sexual_violence_cases__cnmh",
        "V_landmine_cases__cnmh", "V_kidnapping_cases__cnmh",
        "V_child_recruitment_cases__cnmh", "V_conflict_events_all__ucdp",
        "V_homicides_mean__jep", "V_coca_cultivation__aponte",
        "V_farc_presence__aponte", "V_eln_presence__aponte"),
  P = c("P_impl_mean_all_pilars__art", "P_log_dosis_acum__art_inv",
        "P_tratado_acum__art_inv", "P_share_p1_acum__art_inv",
        "P_n_projects_completed__art", "P_aid_usd__apc",
        "P_n_initiatives_total__pdet", "P_n_memory_initiatives__cnmh"),
  J = c("J_n_justice_rooms__slj", "J_slj_present__slj",
        "J_n_cases_processed__casjust", "J_n_conciliadores__concil",
        "J_n_restitution_sentences__urt", "J_n_veedurias__rues",
        "J_land_disputes_total__fiscalia", "J_cdj_cases__minjusticia"),
  S = c("S_nbi__dnp", "S_ipm__dnp", "S_gdp_percapita_constant__dnp",
        "S_gini_land__igac", "S_pop_total__dnp",
        "S_rurality_index__dnp", "S_property_tax_pc__dnp",
        "S_transfers_sgp_pc__dnp"),
  G = c("G_terrain_ruggedness__prio", "G_nightlight__prio",
        "G_travel_time_mean__prio", "G_forest_share__prio",
        "G_indigenous_pct_covered__ant", "G_afro_pct_covered__ant",
        "G_peasant_pct_covered__ant")
)

# ── Static ggplot2 atlas maps (avoids embedding large GeoJSON in HTML) ──
# Each category produces a faceted PNG -- one panel per variable.
# Leaflet is reserved for single-layer maps (missingness, LISA) where
# interactivity matters and file size is controlled.
build_atlas_map <- function(cat_key, viridis_option = "C") {
  cat_vars <- intersect(ATLAS_VARS[[cat_key]], names(geo_cross))
  cat_vars <- cat_vars[sapply(cat_vars, function(v) sum(!is.na(geo_cross[[v]])) >= 5)]
  if (length(cat_vars) == 0) {
    cat("> No variables with sufficient data for this category.\n")
    return(invisible(NULL))
  }

  # Build long-format sf: one row per (muni × variable)
  plot_list <- lapply(cat_vars, function(v) {
    label <- gsub("__.*", "", gsub("^[VPJSG]_", "", v))
    geo_cross |>
      dplyr::transmute(
        dane_code,
        muni_name,
        value     = .data[[v]],
        var_label = label
      )
  })
  plot_sf <- dplyr::bind_rows(plot_list) |> sf::st_as_sf()

  # log1p transform for display (handles extreme right skew)
  plot_sf <- plot_sf |>
    dplyr::mutate(value_display = log1p(pmax(value, 0, na.rm = TRUE)))

  ncols <- min(3L, length(cat_vars))

  ggplot2::ggplot(plot_sf) +
    ggplot2::geom_sf(ggplot2::aes(fill = value_display),
                     color = "white", linewidth = 0.15) +
    ggplot2::scale_fill_viridis_c(
      option   = viridis_option,
      na.value = "#e0e0e0",
      name     = "log(1+x)",
      guide    = ggplot2::guide_colorbar(barwidth = 6, barheight = 0.5,
                                         title.position = "top")
    ) +
    ggplot2::facet_wrap(~var_label, ncol = ncols) +
    ggplot2::theme_void(base_size = 9) +
    ggplot2::theme(
      legend.position  = "bottom",
      strip.text       = ggplot2::element_text(size = 7, face = "bold",
                                                margin = ggplot2::margin(b = 3)),
      plot.margin      = ggplot2::margin(4, 4, 4, 4)
    )
}
```

## How to read this panel

Each tab groups variables by **category**. Use the **layer control** (top-right of the map) to switch between variables within each category.

- **Choropleth maps** show the **municipality mean** across all available years. Values are averaged over each variable's full temporal range (2000-2025 for most sources).
- Click any municipality to see its name and DANE code in a popup.
- Gray polygons have no data for that variable.
- **Moran's I** is computed using Queen contiguity weights restricted to the 170 PDET municipalities.

::: {.callout-note}
**Color scales:** Violence = red/orange, Peacebuilding = green, Justice = blue, Socioeconomic = brown, Geographic = purple-green. Darker = higher value.
:::

::: {.callout-tip collapse="true"}
## What each category captures

**Violence & Conflict (V_):** Documented conflict events from multiple independent sources -- CNMH (historical memory), UCDP (international battle events), VERDATA/JEP (validated homicide estimates). Because each source has different geographic and temporal coverage, disagreements between layers are informative, not errors.

**Peacebuilding & Implementation (P_):** State-led post-agreement programs. SIIPO/ART tracks implementation progress of the PDET development framework across 8 pillars (rural reform, illicit crops, infrastructure, etc.). APC tracks international aid flows. Note: most implementation data only exists from 2017 onward.

**Justice & Dispute Resolution (J_):** Formal and alternative dispute resolution infrastructure. SLJ = community justice centers co-locating mediation, legal aid, and psychosocial services. Conciliadores = certified community mediators. URT = land restitution tribunal outcomes. These are the dissertation's primary outcome variables.

**Socioeconomic & Structural (S_):** Municipal fiscal capacity, population, poverty, and land inequality from Colombia's DNP (national planning). The land Gini (IGAC) measures land concentration -- PDET municipalities average 0.73, well above the national threshold for high inequality.

**Geographic & Territorial (G_):** Terrain ruggedness, nighttime light intensity (proxy for economic activity and state presence), forest/land cover, and the extent of recognized indigenous, Afro-Colombian, and peasant reserve territories. These variables are mostly structural (time-invariant).
:::

---

## Variable Atlas

::: {.panel-tabset}

### Violence & Conflict

::: {.callout-note appearance="minimal"}
Six representative violence indicators. Use the layer switcher (top-right) to compare sources. Selective killings and UCDP events tend to cluster in the same subregions (Catatumbo, Bajo Cauca, Pacific coast) -- where they diverge, it reflects different counting methodologies rather than measurement error.
:::

```{r violence-map}
#| fig-height: 10
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis)) build_atlas_map("V", "C")
```

### Peacebuilding

::: {.callout-note appearance="minimal"}
Implementation progress and aid flows. SIIPO implementation index ranges 0-1 (0 = no implementation, 1 = complete). Investment figures are in Colombian pesos. APC aid is in USD. Note: all ART/SIIPO variables are post-2016 only -- municipalities appear unimplemented before the agreement by construction, not by absence.
:::

```{r peace-map}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis)) build_atlas_map("P", "D")
```

### Justice & Dispute

::: {.callout-note appearance="minimal"}
Dispute resolution infrastructure. SLJ justice rooms (0 = none, 1-2 = present). Conciliadores are certified community mediators registered with the Ministry of Justice. High spatial concentration: justice infrastructure is far from uniformly distributed even within PDET municipalities.
:::

```{r justice-map}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis)) build_atlas_map("J", "B")
```

### Socioeconomic

::: {.callout-note appearance="minimal"}
Structural socioeconomic conditions. Per-capita income and spending are in Colombian pesos (current). Land Gini (IGAC) ranges 0-1, where values above 0.7 indicate extreme land concentration -- the norm in PDET municipalities.
:::

```{r socio-map}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis)) build_atlas_map("S", "A")
```

### Geographic

::: {.callout-note appearance="minimal"}
Structural geographic and territorial conditions. Terrain ruggedness (higher = harder to reach). Nightlight intensity is a proxy for economic activity and state presence: low values dominate across PDET territory. Indigenous/Afro/peasant territorial coverage reflects autonomous governance structures that mediate state presence.
:::

```{r geo-map}
#| fig-height: 9
#| fig-width: 9
#| out-width: "100%"
if (!is.null(pdet_munis)) build_atlas_map("G", "E")
```

:::

---

## Global Moran's I Summary Table

::: {.callout-note appearance="minimal"}
**How to read this table:** Moran's I ranges from -1 (perfect dispersion) to +1 (perfect clustering). Values near 0 mean the variable is spatially random. A positive, statistically significant I means nearby municipalities tend to have similar values -- which is expected for conflict and governance indicators that spread through contiguous territory. The test uses Queen contiguity weights restricted to the 170 PDET municipalities.
:::

```{r morans-table}
if (!is.null(pdet_munis) && nrow(panel) > 0) {
  # Limit to key variables to avoid memory overflow
  morans_vars <- intersect(
    c(unlist(ATLAS_VARS), grep("^J_", numeric_vars, value = TRUE)),
    numeric_vars
  )
  morans_rows <- purrr::map_dfr(morans_vars, function(v) {
    vals <- geo_cross[[v]]
    if (sum(!is.na(vals)) < 30) return(NULL)
    vals[is.na(vals)] <- 0
    if (sd(vals) == 0) return(NULL)
    mi <- compute_morans(vals)
    category <- dplyr::case_when(
      startsWith(v, "V_") ~ "Violence & Conflict",
      startsWith(v, "P_") ~ "Peacebuilding",
      startsWith(v, "J_") ~ "Justice & Dispute",
      startsWith(v, "S_") ~ "Socioeconomic",
      startsWith(v, "G_") ~ "Geographic",
      TRUE ~ "Other"
    )
    tibble::tibble(
      Variable    = v,
      Category    = category,
      Source      = sub(".*__", "", v),
      `Moran's I` = round(mi["I"], 4),
      `p-value`   = round(mi["p"], 5),
      Significant = if (!is.na(mi["p"])) mi["p"] < 0.05 else NA
    )
  })

  if (nrow(morans_rows) > 0) {
    DT::datatable(
      morans_rows |> dplyr::arrange(dplyr::desc(`Moran's I`)),
      options  = list(pageLength = 25, dom = "ftp"),
      rownames = FALSE,
      filter   = "top",
      caption  = "Global Moran's I — Queen contiguity, PDET municipalities (n=170), mean across years."
    ) |>
      DT::formatStyle(
        "Significant",
        backgroundColor = DT::styleEqual(
          c(TRUE, FALSE, NA), c("#d4edda", "#f8d7da", "#f5f5f5")
        )
      )
  }
} else {
  cat("> Panel or geometry not loaded.\n")
}
```

---

## Missingness Map

```{r missingness-map}
if (!is.null(pdet_munis) && nrow(panel) > 0) {
  # Compute per-municipality observability from panel
  na_mat <- is.na(panel[, numeric_vars])
  miss_summary <- panel |>
    dplyr::select(dane_code) |>
    dplyr::mutate(n_obs = rowSums(!na_mat),
                  pct_obs = round(100 * n_obs / length(numeric_vars), 1)) |>
    dplyr::group_by(dane_code) |>
    dplyr::summarise(pct_observed = round(mean(pct_obs), 1), .groups = "drop")

  miss_map <- pdet_munis |>
    dplyr::left_join(miss_summary, by = "dane_code")

  pal_miss <- leaflet::colorNumeric("RdYlGn", domain = c(0, 100), na.color = "#ccc")

  leaflet::leaflet(miss_map) |>
    leaflet::addProviderTiles("CartoDB.Positron") |>
    leaflet::setView(lng = -73.5, lat = 4.5, zoom = 6) |>
    leaflet::addPolygons(
      fillColor   = ~pal_miss(pct_observed),
      fillOpacity = 0.7,
      weight      = 0.5,
      color       = "white",
      popup       = ~paste0("<b>", muni_name, "</b> (", dane_code, ")<br>",
                            "Observed: ", pct_observed, "%")
    ) |>
    leaflet::addLegend(pal = pal_miss, values = c(0, 100),
                       title = "% non-NA", position = "bottomright")
} else {
  cat("> Panel or geometry not loaded.\n")
}
```

::: {.callout-warning}
**Systematic missingness in remote municipalities is a substantive finding.** High missingness in remote areas likely reflects state absence — one of the variables your dissertation theorizes about.
:::
